---
title: "Development Analysis"
author: "Peter Finnerty"
date: "`r format(Sys.Date())`"
output:
  html_document:
    keep_md: true
---


This is an exploration of several Development focused Data sets.


```{r results='hide', message=FALSE, warning=FALSE}
library(ggplot2)
library(tidyverse)
library(ggrepel)
library(kableExtra)
```

####Oxford Multidimensional Poverty Datasets 

```{r}
setwd('/Users/peterfinnerty/R Documents/Independent/Development_Projects/Datasets/Oxford_Dev')
national <- read.csv('MPI_national.csv')
subnational <- read.csv('MPI_subnational.csv')
```


MPI scores measure poverty on multiple dimensions for each country. They allow inter and intracountry comparisons of poverty, capturing more information than simple PPP thresholds.

###Top Ten Countries by Rural Poverty:
```{r results = 'asis'}
national %>% 
  mutate(Rural.Rank = min_rank(-MPI.Rural)) %>% 
  select(Rural.Rank, Country, MPI.Rural) %>%
  arrange(desc(MPI.Rural)) %>% 
  head(10) %>% 
  kable() %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, position = "left")
```

###Top Ten Countries by Urban Poverty:
```{r results = 'asis'}
national %>% 
  mutate(Rural.Rank = min_rank(-MPI.Urban)) %>% 
  select(Rural.Rank, Country, MPI.Urban) %>%
  arrange(desc(MPI.Urban)) %>% 
  head(10) %>% 
  kable() %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, position = "left")
```

###Countries with the Worst Regional Differences:
```{r results = 'asis'}
max <- subnational %>% 
  group_by(Country) %>%
  filter(MPI.Regional == max(MPI.Regional)) %>%
  select(ISO.country.code, Country, Sub.national.region, MPI.Regional, MPI.National) %>% 
  rename(Highest.Region = Sub.national.region, Highest.MPI = MPI.Regional)
min <- subnational %>% 
  group_by(Country) %>%
  filter(MPI.Regional == min(MPI.Regional)) %>%
  select(ISO.country.code, Country, Sub.national.region, MPI.Regional, MPI.National) %>% 
  rename(Lowest.Region = Sub.national.region, Lowest.MPI = MPI.Regional)
gaps <- merge(max, min) %>% 
  mutate(MPI.Gap = Highest.MPI - Lowest.MPI)

gaps %>%
  mutate(Rank = min_rank(-MPI.Gap)) %>% 
  select(Rank, Country, Highest.Region, Lowest.Region, MPI.Gap) %>%
  arrange(Rank) %>% 
  head(10) %>% 
  kable() %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = FALSE, position = "left")
```

Let's look add more country level information.

####World Bank Income Data:
```{r}
setwd('/Users/peterfinnerty/R Documents/Independent/Development_Projects/Datasets/WB/WB_Development_Indicators')
Country_data <- read.csv('WDICountry.csv')
WDI <- read.csv('2014_WDI_Clean.csv')
df <- merge(gaps, WDI, by = 'ISO.country.code')
```

###National MPI vs. GNI Per Capita:
```{r message=FALSE, warning=FALSE}
df %>% 
  ggplot(aes(x = MPI.National, y = GNI_Per_Capita_PPP_2011_Constant)) +
  geom_point() +
  geom_smooth(se = FALSE)
```

We can see that countries with higher incomes experience lower average MPI Scores.

Let's look at countries where this doesn't hold. We need to find nations with high MPIs and high GNI per capita.

###High/Extreme MPIs, ranked by their GNIs:
```{r}
df %>% 
  drop_na(GNI_Per_Capita_PPP_2011_Constant) %>% 
  filter(MPI.National > .3 | Highest.MPI > .5) %>% 
  mutate(Rank = min_rank(-GNI_Per_Capita_PPP_2011_Constant)) %>% 
  arrange(desc(GNI_Per_Capita_PPP_2011_Constant)) %>% 
  select(Rank, Country, GNI_Per_Capita_PPP_2011_Constant, MPI.Gap, MPI.National, 
         Infant_Mortality_per_1000, Life_Expectancy_Birth) %>% 
  kable() %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, position = "left")
```

We can see that Nigeria has the highest GNI, yet has an Extreme MPI Gap.

Let's see look at Nigeria in detail.

###MPI by region in Nigeria:
```{r}
subnational %>% 
  filter(Country == 'Nigeria') %>%
  ggplot() + geom_bar(mapping = aes(x = reorder(Sub.national.region, -MPI.Regional), y = MPI.Regional), stat = 'identity') +
  theme(axis.text.x = element_text(angle = 90, hjust = 1), axis.title.x=element_blank())
```

Unsurprisingly, agricultural regions account for the highest MPIs and Lagos has the lowest MPI score.

Nigeria is an interesting example, as Lagos is one of the most developed cities in the Africa. 
The comparison between max MPI and min MPI may unfairly punish nations with economically strong cities.

To test this, let's see where Lagos ranks in terms of all MPI regions in SSA.

###SSA Regions by MPI:
```{r}

sub <- merge(subnational, Country_data[, c("Country.Code", "Income.Group")], 
             by.x='ISO.country.code', by.y = 'Country.Code')
sub %>% 
  filter(World.region == 'Sub-Saharan Africa') %>%
  mutate(Rank = min_rank(MPI.Regional)) %>% 
  arrange(Rank) %>% 
  select(Rank, Sub.national.region, Country, MPI.Regional, MPI.National) %>%
  filter(Rank <= 10) %>% 
  kable() %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, position = "left")
```

As expected, Lagos has one of the lowest MPI scores in all of SSA. Our previous comparison penalizes Nigeria for this economic prosperity.

We can handle this by comparing the maximum MPI region to the national average.

####Max MPI to National Average:

```{R}
gaps <- gaps %>% mutate(MPI.Gap.Avg = MPI.National - Lowest.MPI)
df <- merge(gaps, WDI, by = 'ISO.country.code')
df %>% 
  drop_na(GNI_Per_Capita_PPP_2011_Constant) %>% 
  filter(MPI.National > .3 | Highest.MPI > .5) %>% 
  mutate(Rank = min_rank(-MPI.Gap.Avg)) %>% 
  arrange(Rank) %>% 
  select(Rank, Country, MPI.Gap.Avg, Highest.MPI, Highest.Region, 
         MPI.National, Lowest.MPI, Lowest.Region, MPI.Gap) %>% 
  kable() %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, position = "left")
```

Here we can see that the impact of Lagos on Nigeria's rank is minimized. 
Additionally, Ethiopia and Mozambique stand out as having extremely high MPI regions versus the average, indicating extreme disparity between those living in Addia Ababa/Maputo and those not.

Let's examine the regions of Ethiopia and Mozambique to see this:

####Ethiopia Regions:
```{r}
subnational %>% 
  filter(Country == 'Ethiopia') %>%
  ggplot() + geom_bar(mapping = aes(x = reorder(Sub.national.region, -MPI.Regional), y = MPI.Regional), stat = 'identity') +
  theme(axis.text.x = element_text(angle = 90, hjust = 1), axis.title.x=element_blank())
```

####Mozambique Regions:
```{r}
subnational %>% 
  filter(Country == 'Mozambique') %>%
  ggplot() + geom_bar(mapping = aes(x = reorder(Sub.national.region, -MPI.Regional), y = MPI.Regional), stat = 'identity') +
  theme(axis.text.x = element_text(angle = 90, hjust = 1), axis.title.x=element_blank())
```

You can see the disparity is drastic in these countries. 


To get a full picture of disparity, let's look at countries where the difference between the 1st and 2nd lowest MPIs is the greatest.

This jump should highlight countries where wealth and development is highly concentrated.

####Ranked by MPI Jump:
```{r}
sub <- subnational %>% 
  group_by(Country) %>% 
  mutate(ranks = order(order(MPI.Regional))) %>% 
  arrange(Country, ranks) %>% 
  mutate(lag_MPI = MPI.Regional - lag(MPI.Regional)) %>% 
  filter(ranks == 2)
df <- merge(df, sub[, c('ISO.country.code', 'lag_MPI')], by = 'ISO.country.code') %>% 
  rename(MPI.Jump = lag_MPI)
df %>% 
  drop_na(GNI_Per_Capita_PPP_2011_Constant) %>% 
  filter(MPI.National > .3 | Highest.MPI > .5) %>% 
  mutate(Rank = min_rank(-MPI.Jump)) %>% 
  arrange(Rank) %>% 
  select(Rank, Country, MPI.Jump, MPI.Gap.Avg, Highest.MPI, Highest.Region, 
         MPI.National, Lowest.MPI, Lowest.Region) %>% 
  head(15) %>% 
  kable() %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, position = "left")
```

The concentration of poverty in these countries is the highest outside of privileged areas.
